---
title: "A Decade of Rent Burden"
author: "Andrew J. Van Leuven"
date: "2026-09-08"
categories: [map]
description: "How the share of rent-burdened households changed between the 2010-2014 and 2020-2024 ACS."
format:
  html:
    theme: cosmo
    embed-resources: false
execute:
  echo: false
  warning: false
  message: false
---

A household is conventionally called **rent burdened** when gross rent eats up 30 percent or more of household income, a threshold codified by the Brooke Amendment and used by HUD, the Census Bureau, and Harvard's Joint Center for Housing Studies. This post takes that measure from ACS table `B25070` at two points a decade apart, the 2010-2014 five-year estimates and the 2020-2024 five-year estimates, and maps the change.

The short answer is that the burdened share fell almost everywhere, and that the first map you would draw of this is largely a picture of sampling error rather than of housing markets. Getting from one to the other is most of what this post is about.

Everything below runs from this one file. The complete script is at the bottom of the post.

```{r}
#| label: setup

library(tidyverse)
library(tidycensus)
library(tigris)
library(sf)
library(spdep)
library(scales)

options(tigris_use_cache = TRUE)
```

## Pulling the data

`B25070` breaks renter households into bins by the share of income they spend on gross rent. Anything from `_007` up is 30 percent or more. Line `_001` is the total, but it includes `_011`, the households whose ratio could not be computed (zero or negative income, no cash rent), so that group comes out of the denominator.

Margins of error are carried through the whole calculation, because they turn out to be the entire story.

```{r}
#| label: pull-acs

VARS <- c("B25070_001", "B25070_007", "B25070_008",
          "B25070_009", "B25070_010", "B25070_011")

# Standard Census formulas for derived estimates: root-sum-square for sums and differences, and the proportion formula for a share. Written out so they stay vectorized over the whole data frame.
moe_rss <- function(...) sqrt(rowSums(cbind(...)^2))
moe_share <- function(num, den, m_num, m_den) {
  p <- num / den
  under <- m_num^2 - (p^2 * m_den^2)
  # Negative radicand: fall back to the ratio formula, per Census guidance
  sqrt(ifelse(under < 0, m_num^2 + (p^2 * m_den^2), under)) / den
}

pull_burden <- function(year, geometry = FALSE) {
  get_acs(
    geography = "county", variables = VARS, year = year, survey = "acs5",
    output = "wide", geometry = geometry,
    resolution = "20m",   # generalized boundaries: ~50x smaller, plots faster
    cache_table = TRUE
  ) |>
    rename_with(\(x) str_replace(x, "^B25070_", "v"), starts_with("B25070_")) |>
    mutate(
      renters      = v001E - v011E,
      renters      = if_else(renters > 0, renters, NA_real_),
      renters_moe  = moe_rss(v001M, v011M),
      burdened     = v007E + v008E + v009E + v010E,
      burdened_moe = moe_rss(v007M, v008M, v009M, v010M),
      pct          = 100 * burdened / renters,
      pct_moe      = 100 * moe_share(burdened, renters, burdened_moe, renters_moe),
      pct_severe   = 100 * v010E / renters
    ) |>
    select(GEOID, NAME, any_of("geometry"),
           renters, burdened, pct, pct_moe, pct_severe)
}

b14 <- pull_burden(2014)
b24 <- pull_burden(2024, geometry = TRUE)
```

Joining the two vintages on GEOID drops a handful of counties whose codes changed in between: Connecticut replaced its eight counties with nine planning regions in 2022, Alaska split Valdez-Cordova, and South Dakota renamed Shannon County.

```{r}
#| label: build-change

burden <- b24 |>
  select(GEOID, NAME, renters_24 = renters, burdened_24 = burdened,
         pct_24 = pct, pct_24_moe = pct_moe, pct_severe_24 = pct_severe) |>
  separate_wider_delim(NAME, ", ", names = c("county", "state"),
                       too_many = "merge") |>
  inner_join(
    b14 |> select(GEOID, renters_14 = renters, burdened_14 = burdened,
                  pct_14 = pct, pct_14_moe = pct_moe, pct_severe_14 = pct_severe),
    by = "GEOID"
  ) |>
  st_as_sf() |>
  mutate(
    change     = pct_24 - pct_14,
    change_moe = moe_rss(pct_24_moe, pct_14_moe),
    sig        = abs(change) > change_moe,   # 90% CI excludes zero
    change_severe = pct_severe_24 - pct_severe_14
  ) |>
  filter(!is.na(change)) |>
  shift_geometry()   # AK/HI insets

states_sf <- burden |> group_by(state) |> summarise(.groups = "drop")
```

Across `r format(nrow(burden), big.mark = ",")` counties, the median change was **`r sprintf("%+.1f", median(burden$change))` percentage points**. Weighting by renter households, the national burdened share went from **`r sprintf("%.1f", 100*sum(burden$burdened_14)/sum(burden$renters_14))`%** to **`r sprintf("%.1f", 100*sum(burden$burdened_24)/sum(burden$renters_24))`%**.

## The raw change

```{r}
#| label: map-helper

# Both maps are the same diverging choropleth, differing only in the column they fill by and how wide the scale runs.
burden_map <- function(col, cap, title, subtitle) {
  ggplot(burden) +
    geom_sf(aes(fill = .data[[col]]), color = NA) +
    geom_sf(data = states_sf, fill = NA, color = "white", linewidth = 0.2) +
    scale_fill_gradient2(
      low = "#2166ac", mid = "#f7f7f7", high = "#b2182b", midpoint = 0,
      limits = c(-cap, cap), oob = scales::squish,
      labels = \(x) sprintf("%+g", x),
      name = "Change in share of\nrenters paying 30%+\nof income (pp)",
      guide = guide_colorbar(barheight = 10, barwidth = 0.8)
    ) +
    labs(title = title, subtitle = subtitle,
         caption = "Source: ACS 5-year estimates, table B25070. AK, HI and PR shifted and rescaled; CT omitted (2022 planning-region change).") +
    theme_void(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 15),
      plot.subtitle = element_text(color = "grey30", margin = margin(b = 8)),
      plot.caption = element_text(color = "grey50", hjust = 0),
      legend.title = element_text(size = 9, lineheight = 1.1),
      plot.background = element_rect(fill = "white", color = NA)
    )
}
```

```{r}
#| label: map-raw
#| fig-width: 10
#| fig-height: 6.5
#| dpi: 120

burden_map("change", 15,
  "Rent burden fell across most of the country",
  "Raw change in the share of renter households paying 30%+ of income, 2010-2014 to 2020-2024 ACS")
```

That map is mostly wrong, and it is worth saying why before going any further.

The ACS is a sample, and in a county with a few hundred renter households the margin of error on a decade-over-decade change runs past ±18 points. Only **`r sprintf("%.0f", 100*mean(burden$sig))`%** of counties have a change a 90 percent confidence interval can tell apart from zero.

## Shrinking out the noise

Rather than blanking those counties out, shrink them. Comparing the spread of the estimates against their own margins of error gives an estimate of how much the true county changes actually vary; each county's value is then blended with a prior in proportion to how precise it is. A well-measured county keeps its own number, a hopeless one leans on the prior, and every county still appears on the map.

The prior matters more than the machinery. The obvious choice is the county's state average, and it is a bad one. Precision-weighting means a state's average is really just its largest metro: Clark County holds 77 percent of Nevada's weight, Maricopa 64 percent of Arizona's. Shrinking toward that paints Las Vegas across a hundred thousand square miles of rural Nevada and invents hard discontinuities at state lines. So the prior here is each county's **neighbors** instead, which is what disease mapping has done since Besag, York and Mollié. Nye County borrows from Clark because Pahrump really is a Las Vegas exurb; Eureka County, three hundred miles north, does not.

```{r}
#| label: shrinkage

# Queen contiguity. The seven island counties (Nantucket, Kauai, Vieques and friends) have no land neighbors, so they borrow from their five nearest by centroid instead.
nb  <- poly2nb(burden, queen = TRUE, snap = 1000)
knn <- knn2nb(knearneigh(st_centroid(st_geometry(burden)), k = 5))
for (i in which(card(nb) == 0)) nb[[i]] <- knn[[i]]

# Observed change y_i = truth + sampling error, with SE = MOE/1.645 (ACS margins are 90%). The prior for each county is the precision-weighted mean of its neighbors' changes, so a well-measured neighbor counts for more than a noisy one.
burden <- burden |> mutate(se = change_moe / 1.645, w = 1 / pmax(se, 1e-6)^2)
d <- st_drop_geometry(burden)
nbr_prior <- map_dbl(nb, \(nn) sum(d$w[nn] * d$change[nn]) / sum(d$w[nn]))

# Between-county variance tau^2 by Paule-Mandel: the value that makes the chi-square residual around the prior come out at its expectation. It sets how much borrowing happens, so no cutoff is chosen by hand.
s2 <- d$se^2
gap <- function(t2) sum((d$change - nbr_prior)^2 / (s2 + t2)) - (nrow(d) - 1)
tau2 <- if (gap(0) <= 0) 0 else uniroot(gap, c(0, 1e4))$root

burden <- burden |>
  mutate(
    nbr_prior   = nbr_prior,
    B           = s2 / (s2 + tau2),
    change_eb   = (1 - B) * change + B * nbr_prior,
    reliability = 1 - B   # 0 = all borrowed, 1 = all own data
  )
```

The between-county standard deviation of the *true* change comes out to **`r sprintf("%.1f", sqrt(tau2))` points**, against an observed spread of **`r sprintf("%.1f", sd(burden$change))`**. Roughly **`r sprintf("%.0f", 100*(1 - tau2/var(burden$change)))`%** of the variation in the first map is sampling noise. Note the narrower scale below.

```{r}
#| label: map-eb
#| fig-width: 10
#| fig-height: 6.5
#| dpi: 120

burden_map("change_eb", 8,
  "Rent burden eased most in the eastern half of the country",
  "Change in the share of renter households paying 30%+ of income, shrunk toward neighboring counties")
```

The median county's shrunk value takes **`r sprintf("%.0f", 100*median(burden$reliability))`%** of its weight from its own data, rising to **`r sprintf("%.0f", 100*median(burden$reliability[burden$renters_24 > 20000]))`%** among the largest counties.

## Notes

A few things worth stating plainly before anyone quotes a number off these maps.

**The 30 percent threshold is a policy artifact, not a finding.** It descends from an old rule of thumb (a week's wages for a month's rent) that the Brooke Amendment codified at 25 percent in 1969, and that Congress raised to 30 percent in 1981 as a budget measure. The competing approach is residual income, which asks whether what remains after rent covers food, transportation, and health care. A ratio treats a household earning $30,000 and one earning $300,000 as equally burdened at the same percentage, which they plainly are not.

**A share defined by a threshold hides intensity.** Everyone moving from 29 to 31 percent looks like a crisis. Everyone moving from 35 to 49 percent looks like nothing happened. `B25070` reports only bins, so the median rent-to-income ratio cannot be recovered from it.

**The denominator moved for reasons unrelated to rent.** The 2020-2024 window covers the stimulus payments, the expanded child tax credit, supplemental unemployment insurance, and emergency rental assistance, all of which raised measured household income. The 2010-2014 window sits near the post-foreclosure peak of renter cost burden. Some of the national decline is the income side of the ratio rather than rents becoming affordable.

**Falling burden can mean displacement.** The measure conditions on who still lives in the county. If cost-burdened renters get priced out, the county's rate improves.

**Shrinkage is conservative by construction.** A county that genuinely diverged from its neighbors gets pulled back toward them, so real local outliers are understated. That is the standard small-area-estimation bargain: a little bias bought in exchange for a map that is legible. The technique is ordinary (the Census Bureau's own SAIPE and SAHIE programs are built on the closely related Fay-Herriot model), but the choice of prior does real work, and a neighbor prior is an assumption about geography rather than a neutral one.

**The prior is treated as known.** Each county's neighbor average is itself estimated from noisy data, and the variance calculation here ignores that. The effect is to overstate tau slightly and therefore to shrink slightly less than a fully specified spatial model would.

**Geography.** Connecticut is absent because it replaced its eight counties with nine planning regions in 2022, leaving no GEOID overlap between the two vintages. Alaska's Valdez-Cordova split into Chugach and Copper River, Wade Hampton became Kusilvak, and Shannon County, South Dakota became Oglala Lakota County. Puerto Rico's 78 municipios are included.

**Source.** ACS 5-year estimates, table B25070, via `tidycensus`. Geometry from `tigris` at 20m generalized resolution.

## Data

```{r}
#| label: download
#| results: asis

# Encode the tidied table straight into the page as a data URI, so the post stays a single file with no data sidecars in the repo.
csv <- burden |>
  st_drop_geometry() |>
  select(GEOID, county, state, renters_14, renters_24,
         pct_14, pct_24, change, change_moe, change_eb, reliability, sig) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  format_csv()

cat(sprintf(
  '<a href="data:text/csv;base64,%s" download="rent_burden.csv" class="btn btn-primary">Download CSV</a>',
  jsonlite::base64_enc(charToRaw(csv))
))
```

```{r}
#| label: fullscript
#| results: asis

# Rebuilds the post's code as one script by reading this file back and stitching the chunks together, so the listing below can never drift out of sync with what actually ran.
src    <- readLines("index.qmd")
opens  <- grep("^```\\{r", src)
closes <- grep("^```\\s*$", src)

blocks <- map(opens, \(o) {
  body  <- src[(o + 1):(closes[closes > o][1] - 1)]
  label <- str_match(body[1], "#\\|\\s*label:\\s*(\\S+)")[, 2]
  if (identical(label, "fullscript")) return(NULL)
  body <- body[!str_detect(body, "^#\\|")]
  body <- body[cumsum(body != "") > 0]
  c(paste0("# ---- ", label, " ", strrep("-", max(0, 62 - nchar(label)))), body, "")
})

cat('::: {.callout-note collapse="true" title="Full script"}\n\n```r\n')
cat(paste(unlist(compact(blocks)), collapse = "\n"))
cat("\n```\n\n:::\n")
```
