---
title: "Mapping Golf Courses Across the United States"
author:
  - "Andrew J. Van Leuven"
  - "Trey Malone"
date: "2026-07-03"
categories: [map]
description: "A first look at where the country's golf courses are located, by count, per capita, and as point data."
format:
  html:
    theme: cosmo
    embed-resources: false
execute:
  # Pinned. This post embeds its geometry in the mapgl widgets, so each re-execution writes a fresh multi-megabyte result into git history. Flip to auto when actually editing the post.
  freeze: true
---

*Draft shell. No editorializing yet.*

This post maps golf courses across the United States. The underlying point data come from a [POI Factory](http://www.poi-factory.com/node/29395) dataset of golf-course locations, which we cleaned, geocoded to counties, and joined to 2020 Census population counts. The [`prep_golf_data.R`](prep_golf_data.R) script handles all of that and writes the two GeoJSON files used below.

Every map has an **All / Public / Private** toggle in the top-left. "Public" here means publicly accessible play (daily-fee, municipal, and semi-private courses); "Private" means members-only clubs.

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(sf)
library(mapgl)
library(glue)

counties_sf <- st_read("golf_counties.geojson", quiet = TRUE)
points_sf   <- st_read("golf_points.geojson", quiet = TRUE)

# Lower-48 view + dissolved state outlines for the choropleths
lower48   <- filter(counties_sf, !st %in% c("AK", "HI", "PR"))
us48      <- summarise(lower48)
us_states <- lower48 |> group_by(st) |> summarise(.groups = "drop")

# Category colors used throughout
col_public  <- "#2e7d32"  # green
col_private <- "#c1440e"  # burnt orange

# Trimmed sf objects actually embedded in the maps below. Each map gets one
# shared GeoJSON source (added once via add_source), and the All/Public/
# Private layers read from that single source with a `filter` expression,
# instead of embedding the dataset three times over.
counties_map_sf <- counties_sf |>
  select(county, courses, courses_public, courses_private,
         pop, per_100k, per_100k_public, per_100k_private)

points_map_sf <- points_sf |>
  transmute(
    category,
    tooltip = glue(
      "<b>{course}</b><br>{access} &middot; {holes} holes<br>{city_state}"
    )
  )

public_filter  <- list("==", list("get", "category"), "Public")
private_filter <- list("==", list("get", "category"), "Private")

toggle_layers <- c("All" = "All", "Public" = "Public", "Private" = "Private")

# Choropleth toggle helper: one shared source, three fill layers (All/Public/
# Private) each reading a different column, wired to a layers control.
# `interpolate()` clamps values above the top stop to the top color, which
# gives us a per-capita (or per-count) ceiling for free.
choro_toggle <- function(fill_cols, count_cols, values, stops,
                         tick_labels, legend_title, unit_label) {
  labels <- names(fill_cols)
  m <- maplibre(bounds = us48, style = carto_style("positron")) |>
    add_source(id = "counties_src", data = counties_map_sf)
  for (i in seq_along(fill_cols)) {
    m <- add_fill_layer(
      m,
      id = labels[i],
      source = "counties_src",
      fill_color = interpolate(
        column = fill_cols[[i]], values = values,
        stops = stops, na_color = "#eeeeee"
      ),
      fill_opacity = 0.85,
      tooltip = concat(
        "<b>", get_column("county"), "</b><br>",
        get_column(count_cols[[i]]), " courses<br>",
        "pop: ", get_column("pop")
      ),
      visibility = if (i == 1) "visible" else "none"
    )
  }
  m |>
    add_line_layer(id = "state_outlines", source = us_states,
                   line_color = "white", line_width = 0.4) |>
    add_continuous_legend(legend_title, values = tick_labels, colors = stops,
                          position = "bottom-left") |>
    add_layers_control(layers = setNames(labels, labels),
                       position = "top-left", collapsible = FALSE)
}
```

Across the full dataset there are **`r format(nrow(points_sf), big.mark = ",")`** golf courses (`r format(sum(points_sf$category == "Public"), big.mark = ",")` public, `r format(sum(points_sf$category == "Private"), big.mark = ",")` private) spread across **`r format(n_distinct(counties_sf$GEOID[counties_sf$courses > 0]), big.mark = ",")`** counties.

## Count by County

The raw count of golf courses in each county. The color scale tops out at 30 courses (any county above that shows the darkest shade).

```{r}
#| label: count-map
#| echo: false
#| warning: false
#| message: false

count_stops <- RColorBrewer::brewer.pal(3, "YlGn")

choro_toggle(
  fill_cols   = c(All = "courses", Public = "courses_public",
                  Private = "courses_private"),
  count_cols  = c(All = "courses", Public = "courses_public",
                  Private = "courses_private"),
  values      = c(0, 15, 30),
  stops       = count_stops,
  tick_labels = c("0", "15", "30+"),
  legend_title = "Golf courses per county"
)
```

## Courses per Capita

Golf courses per 100,000 residents, using total population from the 2020 Census PL 94-171 redistricting file (via `tidycensus`). Rather than dropping small-population counties, the color scale is **capped at 50 courses per 100k**, so a handful of tiny counties with extreme rates don't blow out the scale.

```{r}
#| label: percapita-map
#| echo: false
#| warning: false
#| message: false

percap_stops <- RColorBrewer::brewer.pal(3, "PuBuGn")

choro_toggle(
  fill_cols   = c(All = "per_100k", Public = "per_100k_public",
                  Private = "per_100k_private"),
  count_cols  = c(All = "courses", Public = "courses_public",
                  Private = "courses_private"),
  values      = c(0, 25, 50),
  stops       = percap_stops,
  tick_labels = c("0", "25", "50+"),
  legend_title = "Courses per 100k residents"
)
```

## Courses as Points

Following [Kyle Walker's mapgl dots post](https://walker-data.com/posts/mapgl-dots/), the same courses are shown three ways: as individual points, as clusters, and as a heatmap.

### Individual points

```{r}
#| label: points-map
#| echo: false
#| warning: false
#| message: false

point_color <- match_expr(
  column = "category", values = c("Public", "Private"),
  stops = c(col_public, col_private)
)

maplibre(bounds = us48, style = carto_style("positron")) |>
  add_source(id = "points_src", data = points_map_sf) |>
  add_circle_layer(
    id = "All", source = "points_src",
    circle_color = point_color, circle_radius = 3, circle_opacity = 0.6,
    circle_stroke_color = "white", circle_stroke_width = 0.3, tooltip = "tooltip"
  ) |>
  add_circle_layer(
    id = "Public", source = "points_src", filter = public_filter, visibility = "none",
    circle_color = col_public, circle_radius = 3, circle_opacity = 0.6,
    circle_stroke_color = "white", circle_stroke_width = 0.3, tooltip = "tooltip"
  ) |>
  add_circle_layer(
    id = "Private", source = "points_src", filter = private_filter, visibility = "none",
    circle_color = col_private, circle_radius = 3, circle_opacity = 0.6,
    circle_stroke_color = "white", circle_stroke_width = 0.3, tooltip = "tooltip"
  ) |>
  add_categorical_legend(
    "Course type", values = c("Public", "Private"),
    colors = c(col_public, col_private), position = "bottom-left",
    circular_patches = TRUE
  ) |>
  add_layers_control(layers = toggle_layers, position = "top-left",
                     collapsible = FALSE)
```

### Clusters

```{r}
#| label: cluster-map
#| echo: false
#| warning: false
#| message: false

# mapgl's cluster_options always builds a fresh source per layer (clustering
# is a source-level property), so unlike the maps above we can't share a
# single source here. We still keep each input as small as possible (only
# the columns each layer actually needs).
pts_all_c     <- points_map_sf
pts_public_c  <- points_map_sf |> filter(category == "Public") |> select(tooltip)
pts_private_c <- points_map_sf |> filter(category == "Private") |> select(tooltip)

maplibre(bounds = us48, style = carto_style("positron")) |>
  add_circle_layer(
    id = "All", source = pts_all_c, tooltip = "tooltip",
    circle_color = point_color, circle_radius = 4, circle_opacity = 0.8,
    cluster_options = cluster_options(
      color_stops = c("#a5d6a7", "#66bb6a", "#2e7d32"),
      radius_stops = c(15, 25, 35), count_stops = c(0, 50, 250)
    )
  ) |>
  add_circle_layer(
    id = "Public", source = pts_public_c, visibility = "none", tooltip = "tooltip",
    circle_color = col_public, circle_radius = 4, circle_opacity = 0.8,
    cluster_options = cluster_options(
      color_stops = c("#a5d6a7", "#66bb6a", "#2e7d32"),
      radius_stops = c(15, 25, 35), count_stops = c(0, 50, 250)
    )
  ) |>
  add_circle_layer(
    id = "Private", source = pts_private_c, visibility = "none", tooltip = "tooltip",
    circle_color = col_private, circle_radius = 4, circle_opacity = 0.8,
    cluster_options = cluster_options(
      color_stops = c("#f6b48c", "#e8814c", "#c1440e"),
      radius_stops = c(15, 25, 35), count_stops = c(0, 25, 100)
    )
  ) |>
  add_layers_control(layers = toggle_layers, position = "top-left",
                     collapsible = FALSE)
```

### Heatmap

```{r}
#| label: heatmap-map
#| echo: false
#| warning: false
#| message: false

# Heatmap has no tooltip, so this source drops that column entirely (the
# heaviest property on points_map_sf) and only keeps category for filtering.
heat_map_sf <- points_map_sf |> select(category)

maplibre(bounds = us48, style = carto_style("dark-matter")) |>
  add_source(id = "heat_src", data = heat_map_sf) |>
  add_heatmap_layer(
    id = "All", source = "heat_src", heatmap_radius = 12, heatmap_opacity = 0.8
  ) |>
  add_heatmap_layer(
    id = "Public", source = "heat_src", filter = public_filter, visibility = "none",
    heatmap_radius = 12, heatmap_opacity = 0.8
  ) |>
  add_heatmap_layer(
    id = "Private", source = "heat_src", filter = private_filter, visibility = "none",
    heatmap_radius = 12, heatmap_opacity = 0.8
  ) |>
  add_layers_control(layers = toggle_layers, position = "top-left",
                     collapsible = FALSE)
```

## Data

```{r}
#| label: download
#| echo: false
#| results: asis

cat('<a href="golf_courses.csv" download class="btn btn-primary">Download CSV</a>\n')
cat('<a href="golf_points.geojson" download class="btn btn-primary" style="margin-left: 4px;">Download GeoJSON</a>\n')
```
